Annotated example of index standardization with VAST

### 2022-08-03 doc/appendix-pcod-VAST-poisson-link.Rmd
Fixes to unused appendix

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

```{r fit, echo = TRUE, results = 'hide'}
 # create folder for saved output:
dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

@@

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, 
  family = delta_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  points(years, sdmTMB_est$b_j[,sdmTMB_pars])
  segments(years, sdmTMB_est$b_j[,sdmTMB_pars] - 2 * sdmTMB_sd$b_j[,sdmTMB_pars], 
    years, sdmTMB_est$b_j[,sdmTMB_pars] + 2 * sdmTMB_sd$b_j[,sdmTMB_pars],
    col = "red")
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

@@

```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r plot-knots, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
```

@@

```{r plot-index, echo = TRUE, out.width="3in"}
if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
}
if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
}
```

We can compare the index we would get using sdmTMB. Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models. Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws. This should be approximately comparable with the bias-corrected index calculated above with VAST.

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
  area = rep(4, nrow(qcs_grid))
)
ind <- get_index(p, bias_correct = FALSE)
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.1) +
  geom_line(alpha = 0.8) + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```

>>>>

settings <- make_settings(
  n_x = 185, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 1), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

```{r fit, echo = TRUE, results = 'hide'}
 # create folder for saved output:
dir.create(paste0(here("doc", "appendix-VAST-pg")), showWarnings = FALSE)

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
if (FALSE) { # to check offset/effort
  set.seed(1)
  pcod$effort <- exp(rnorm(nrow(pcod), mean = 0, sd = 0.2))
}
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST-pg", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST-pg"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

@@

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh,
  offset = log(pcod$effort),
  family = delta_poisson_link_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  if (sdmTMB_pars == 1) {
    points(years, sdmTMB_est$b_j)
    segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
    col = "red")
  } else {
    points(years, sdmTMB_est$b_j2)
       segments(years, sdmTMB_est$b_j2 - 2 * sdmTMB_sd$b_j2, 
    years, sdmTMB_est$b_j2 + 2 * sdmTMB_sd$b_j2,
    col = "red")
  }
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

@@

```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST-pg"), "/")
)
```

@@

```{r plot-knots, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST-pg", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST-pg", "ln_density-predicted.png"))
```

@@

```{r plot-index, echo = TRUE, out.width="3in"}
if (file.exists(here("doc", "appendix-VAST-pg", "Index-Biomass.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST-pg", "Index-Biomass.png"))
}
if (file.exists(here("doc", "appendix-VAST-pg", "Index.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST-pg", "Index.png"))
}
```

We can compare the index we would get using sdmTMB.

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
ind <- get_index(p, bias_correct = FALSE, area = rep(4, nrow(qcs_grid)))
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST-pg/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.1) +
  geom_line(alpha = 0.8) + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```
